Medical Concept Extraction

### 2012-12-16 paper/paper.tex
Adding a sample row so you guys know how to add stuff to the table.
features section complete

\section{Challenges}
There are several unique challenges which arise due to the the nature of the dataset and the approach of leveraging traditional machine learning classifiers for natural language.

\subsection{High Dimensionality}
In order to utilize traditional machine learning classifiers for natural language processing, it is necessary to generate linguistic features for each item that requires classification. The linguistic features which characterize each item often do not have a meaningful strict ordering (e.g. it is not meaningful to say ``muffin'' $>$ ``bread'' if your feature was the word itself). Therefore, it is necessary to represent each feature without a strict ordering as a binary dimension reflecting the presence or absence of that feature.

Using this binary dimension representation results in a high dimensional space (e.g. using a word itself as a feature results in a number of dimensions equal to the size of the vocabulary of the training data). Further, the high dimensional space is exceptionally sparse since each item generally only satisfies one binary dimension for the feature. For non-asymptotic cases found in practice, this generally means that the feature space grows much faster than the amount of data and most of the theoretical guarantees of the traditional machine learning algorithms do not hold.

\subsection{Biased Data}
Many machine learning classifiers assume relatively balanced data with respect to the class labels available. However, in the task of concept extraction it is most often the case that there is a null concept class indicating that the token is part of no relevant concept which is overwhelmingly more abundant than the concepts. The bias of the class labels in the data presents a challenge because a classifier optimized with respect to accuracy will be inclined to simply assign the null concept class for all labels since that will result in relatively high accuracy. Such classifiers will not, however, perform well with respect to more meaningful measures such as precision, recall, or F1-measure.

\subsection{Amount of Data}
As with many statistically-grounded projects, there is often only a limited amount of data available. Within the medical domain this holds particularly true due to the expertise (and subsequent cost) required to provide annotation and the inability to easily merge data from multiple sources -- either due to the differing formats in which it is stored/accessed or various legal restraints in place to protect patients.

Such dearth of training data means that the amount of test data is often much larger than the amount of training data available. Consequently, it is even more difficult to obtain decent results if the models trained even moderately overfit the training data -- a particular risk due to the sparse distribution of high dimensional items discussed previously.

\subsection{Data Usage Agreements}
In addition to limited availability of data, it is often difficult to obtain the data that is relevant. Even for ``open'' data sets such as those provided by the i2b2 National Center for Biomedical Computing, one must sign a data use agreement in order to maintain compliance with the U.S. Health Insurance and Accountability Act (HIPAA) privacy and security rules. The approval process for ``open'' data sets can still take weeks, while access to more controlled data sets can often take months and require security measures such as single physical machine access.

@@

\subsubsection{Word Shape}
In the same spirit of our word length and MIT RegEx features, we decided to look at the high-level shapes of individual words for better generalization. More specifically, we encode words into new strings that give a sense of their capitalization and punctuation patterns as well as approximate length. We tested a variety of open-sourced word shape classification methods including two methods written by Dan Klein, one of which simply classifies words as numeric, strictly upper or lower cased, or mixed, and a second, more fine-grained classifier that attempts to group words into equivalence classes by collapsing sequences of the same time. Additionally classifiers developed by Christopher Manning seek to make less distinctions mid-sequence by sorting words based on length. Each word shape classifier result added an additional dimension to our feature space. By naively including all word shape alternatives as features, we were able to improve our testbed CRF F1 score from 0.82 to 0.85. After further testing, we settled on \begin{verbatim}______\end{verbatim} as the most promising word shape metric because \begin{verbatim}______.\end{verbatim}

@@

\subsection{N-gram Features}
TODO(SAM):

\subsubsection{Previous}

\subsubsection{Next}

@@

Our framework includes a model class that can be trained on certain data with certain features enabled. The model class supports training a libsvm, liblinear, and crfsuite classifiers that can be used for prediction in the future. One of the nice things, about the model class is that it can be serialized and reloaded dynamically. This is a really nice capability because it allows you to train a large number of models on various subsets of the data and using various different features to see which perform the best.

@@

As mentioned previously, the framework's model class has support for training various classifiers: a Support Vector Machine, a CRF, and L2-regularized Support Vector Machine. The implementation of these algorithms is delegates to various third-party libraries described below.

@@

\item {\tt models/<model-name>} - Contains the meta data specific to each model that was trained. The framework saves the model's internal state (the features used, the input test set, etc.) as well as the trained parameters for the libsvm, crfsuite, and liblinear classifiers. This information is used to deserialize and load a model at runtime for prediction tasks.

@@

\item {\tt liblinear} \cite{liblinear}: liblinear is a linear classifier library that has special support multi-class classification. It also provides easy to use training and predicting command line utilities that accept a similar input format as libsvm.

\item {\tt crfsuite} \cite{crfsuite}: crfsuite is an implementation for conditional random fields and includes command line front ends to train a model on data and to predict the labels on an unseen data set. The data format for the input training files is also a sparse matrix representation but is slightly different from the format used by libsvm and liblinear. Our codebase nicely abstracts out these format differences and produces the right format depending on the model in use.

@@

\begin{figure}[t]
\begin{center}
\begin{center}
	\begin{tabular}{|l|l|}
	\hline
	\multicolumn{2}{|c|}{\bf Testing Data Set} \\ \hline
	{\bf Label} & {\bf Count}  \\ \hline
	test & 17237 \\ \hline
	none & 202438 \\ \hline
	 problem & 30276 \\ \hline
	 treatment & 17298 \\ \hline
	\end{tabular}
\end{center}

\begin{center}
	\begin{tabular}{|l|l|}
	\hline
	\multicolumn{2}{|c|}{\bf Training Data Set} \\ \hline
	{\bf Label} & {\bf Count}  \\ \hline
	test & 8492 \\ \hline
	none & 114783 \\ \hline
	 problem & 17326 \\ \hline
	 treatment & 8940 \\ \hline
	\end{tabular}
\end{center}
\end{center}
\caption{The number of each class the in testing and training data sets.}
\label{fig:data_set_counts}
\end{figure}

@@

Recall that each model contains three different classifiers: libsvm, liblinear, and crfsuite. So, in total we attempted training 69 classifiers. Unfortunately, as discussed later in the paper, the libsvm classifiers took very long to train and most were not finished in time. Consequently, their results are not included in this report. Fortunately we were able to train a single libsvm classifier from a model that had all of the features enabled.  So, all in all the results include the output from 47 classifiers (46 liblinear and crfsuite classifiers and 1 libsvm classifier).

@@

We deployed the system on a machine running Ubuntu 11.10 with a QuadCore AMD Opteron processor running a 2.5Ghz and 48GB of ram.

In general training the liblinear and crfsuite classifiers was pretty quick. Each took around an hour to train on the entire training set. However, the libsvm classifier took approximate 6 hours to train on the same data.

In terms of predicting labels for unseen input, all classifiers performed equally well. They each took around 1 hour to label the entire test data set. However, the vast major of time was spent computing features on the input rather than actually running the classifiers.

@@

\begin{figure}[t]
\begin{center}
	\begin{tabular}{|l|l|l|l|}
	\hline
	\multicolumn{4}{|c|}{\bf crfsuite} \\ \hline
	{\bf Label} & {\bf Precision} & {\bf Recall} & {\bf F1} \\ \hline

test & 0.8650 &	0.6857 &	0.7650 \\ \hline
problem & 0.7686 &	0.7492	& 0.7587 \\ \hline
none & 0.9321 & 	0.9637 &	0.9477 \\ \hline
treatment  & 0.7911 &	0.6763 &	0.7292 \\ \hline

	\hline
	Average: & 0.7512 &	 0.8537 &	0.7891 \\
	\hline
	\end{tabular}
\end{center}

\begin{center}
	\begin{tabular}{|l|l|l|l|}
	\hline
	\multicolumn{4}{|c|}{\bf liblinear} \\ \hline
	{\bf Label} & {\bf Precision} & {\bf Recall} & {\bf F1} \\ \hline

test  & 0.7908 & 0.8347 & 	0.8122 \\ \hline
problem  & 0.5615 & 	0.9246 &	0.6987 \\ \hline
none  & 0.9784	& 0.8646 &	0.9180 \\ \hline
treatment  & 0.6740 &	0.7908	& 0.7278 \\ \hline

	\hline
	Average: & 0.8392 &	 0.7687  &	0.8001 \\
	\hline
	\end{tabular}
\end{center}

\begin{center}	
	\begin{tabular}{|l|l|l|l|}
	\hline
	\multicolumn{4}{|c|}{\bf libsvm} \\ \hline
	{\bf Label} & {\bf Precision} & {\bf Recall} & {\bf F1} \\ \hline

     test & 0.8491 &	0.7392	& 0.7903 \\ \hline
  problem & 0.7603 &	0.7692 &	0.7647 \\ \hline
     none & 0.9395	 & 0.9668	& 0.9529 \\ \hline
treatment & 0.8289 &	0.6365 &	0.7201\\ \hline

	\hline
	Averages: & 0.8444 &	 0.7779  & 	0.8070 \\
	\hline
	\end{tabular}
\end{center}

\caption{Results from a model trained on all of the test data with all of the features enabled.}
\label{fig:all_features_results}
\end{figure}

\begin{figure}
\begin{center}

	\begin{tabular}{|l|l|l|}
	\hline
	{\bf Feature Set} & {\bf Classifier} & {\bf F1 Score}  \\ \hline

	\end{tabular}

\end{center}
\caption{A summary of the F1-scores from many different models with different features enabled.}
\label{fig:summary_f1_scores}
\end{figure}

Overall we found that our system did a good job at predicting labels for medical notes. Figure~\ref{fig:all_features_results} shows summary statistics as well as a confusion matrix of a model that was trained with all the features enabled. The results are broken down so we can see the relative performance of each of the individual classifiers. Overall, the system does very well and achieves an average F1-score of {\bf 0.79}.

As mentioned before, we trained a total of 47 classifiers with different features to get an idea for what features were the most important. A summary of the F1-scores of these classifiers is show in Figure~\ref{fig:summary_f1_scores}.

@@

We stopped using libsvms.

@@

Found that libsvm was not worth while to use

@@

We sincerely thank Dr. Robert Berwick  and Geza Kovacs
for their guidance, help, and support. May the Force be with you,
always.

>>>>

\section{Challenges}

\subsection{High Dimensionality}
TK(TJN)

\subsection{Biased Data}
TK(TJN)

\subsection{Amount of Data}
TK(TJN)

Low amounts of train data with high amounts of tests data.

\subsection{Data Usage Agreements}
TODO(SAM):

@@

\subsubsection{Word Shape}
In the same spirit of our word length and MIT RegEx features, we decided to look at the high-level shapes of individual words for better generalization. More specifically, we encode words into new strings that give a sense of their capitalization and punctuation patterns as well as approximate length. We tested a variety of open-sourced word shape classification methods including two methods written by Dan Klein, one of which simply classifies words as numeric, strictly upper or lower cased, or mixed, and a second, more fine-grained classifier that attempts to group words into equivalence classes by collapsing sequences of the same time. Additionally classifiers developed by Christopher Manning seek to make less distinctions mid-sequence by sorting words based on length. Each word shape classifier result added an additional dimension to our feature space. By naively including all word shape alternatives as features, we were able to improve our testbed CRF F1 score from 0.82 to 0.85. After further testing, we settled on TODO as the most promising word shape metric because TODO.

@@

\subsection{N-gram Features}
Although our primary goal is to classify $individual$ words, a single word in a medical record entry is hardly independent from the next.  More specifically, medical concepts can often span more than one word--we may not be interested in detecting these exact spans, but features from certain words provide insight into how to label their neighbors. For example, ambiguous words such as "heart" could precede "rate" or "disease", and making that distinction between test and problem is much easier when we examine the subsequent word in the sentence containing "heart". A neighboring word can serve as an indicator even if it is not part of the same concept span. For example, words like "showed" and its synonyms are non-medical, but tend to follow a test term. Likewise, "rule out" and "history of" tend to precede problems. With this in mind, we observe all features of immediately neighboring words.

\subsubsection{Previous}
We take a basic approach to context features by considering the immediately preceding word for each individual word. Given a word with the feature set previously described, we simply duplicate each feature with the corresponding value for it's previous neighbor and rename these features $previous\_$<NAME> (for example, $previous\_$length).  This serves to double the size of our feature space.
\subsubsection{Next}
Likewise, we duplicate the subsequent word features, which are relabeled as $next\_$<NAME>. At this point, our feature space is tripled, but was already quite large--hence, we decided only to consider surrounding words at a bigram level. Nevertheless, this relatively straightforward step brought out CRF F1 testing score up to 0.98.

@@

Our framework includes a model class that can be trained on certain data with certain features enabled. The model class supports training a libSVM, libLinear, and CRF classifier that can be used for prediction in the future. One of the nice things, about the model class is that it can be serialized and reloaded dynamically. This is a really nice capability because it allows you to train a large number of models on various subsets of the data and using various different features to see which perform the best.

@@

As mentioned previously, the framework's model class has support for training various classifiers: a Support Vector Machine, a CRF, and linear classifier. The implementation of these algorithms is delegates to various third-party libraries described below.

@@

\item {\tt models/<model-name>} - Contains the meta data specific to each model that was trained. The framework saves the model's internal state (the features used, the input test set, etc.) as well as the trained parameters for the libSVM, crfsuite, and libLinear classifiers. This information is used to deserialize and load a model at runtime for prediction tasks.

@@

\item {\tt liblinear} \cite{liblinear}: lib linear is a linear classifier library that has special support multi-class classification. It also provides easy to use training and predicting command line utilities that accept a similar input format as libsvm.

\item {\tt crfsuite} \cite{crfsuite}: crfsuite is an implementation for conditional random fields and includes command line front ends to train a model on data and to predict the labels on an unseen data set. The data format for the input training files is also a sparse matrix representation but is slightly different from the format used by libsvm and lib linear. Our codebase nicely abstracts out these format differences and produces the right format depending on the model in use.

@@

@@

@@

How long each model took to train and how fast we can predict. Also discussion the hardware that was used to train and predict in terms of CPU speed, memory, etc.

@@

A table of a bunch of awesome results. We should present results by model type (SVM, CRF, LIN) and feature set.

@@

We stopped using libSVMs.

@@

Found that libSVM was not worth while to use

@@

We sincerely thank Dr. Robert Berwick  and Geza Kovacs
for their guidance, help, and support.
